Wait-free Queues with Polylogarithmic Step Complexity

### 2022-08-31 thesis/thesis.tex
review DSearch

\end{multicols}
\end{algorithmic}
\end{algorithm}
\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
\cmt{Returns \tt{<b,i>} if $E_e(root)=E_i(root,b)$.}
\State \tt{start= end-1}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start= max(start-(end-start), 0)} \label{doubling}
\EndWhile
\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
\State\Return \tt{<b,i>}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

\begin{lemma}[DSearch correctness] \label{dsearch}
Assume $\nf{root.blocks[end].sum\sub{enq}}\geq\nf{e}$ and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}.  \nf{DSearch(e, end)} returns \nf{<b, i>} such that $E_i(root,b)= E_e(root)$.
\end{lemma}
\begin{proof}
It is trivial to see that the doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} will find $E_e(root)$ eventually. Because \nf{root.blocks[].sum\sub{enq}} is an increasing value from 0 to some value greater than \nf{e}. So there is a \nf{b} that $\nf{root.blocks[b].sum\sub{enq}}>e$ but $\nf{root.blocks[b-1].sum\sub{enq}}<e$.

First we show $\nf{end}-\nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1\big)$. From line \ref{addOP}, we know that size of the every block in the tree is greater than 0. So each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue} or at least one \nf{Dequeue}. Suppose there were more than \nf{root.blocks[b].\size}\nf{Dequeue}s in \nf{root.blocks[b+1..end-1]}. Then the  queue would become empty at some point after \nf{blocks[b]}'s last operations and before \nf{root.blocks[end]}'s first operation. Which means the response to to a \nf{Dequeue} in \nf{root.blocks[end]} could not be in $E(n,b)$. Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end].\size} \nf{Enqueue}s. Because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size is \nf{root.blocks[end].\size}. Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}. So $\nf{end-b}\leq 2 \times\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1$. See Figure \ref{fig::doubling}.

Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]} then with doubling search in $\Theta \big(\log($\nf{root.blocks[b].\size+root.blocks[end].\size}$)\big)$ steps we reach \nf{start=c} that the \nf{root.blocks[c].sum\sub{enq}} is less than \nf{e} and \nf{end-c} is not more than $2\times$\nf{root.blocks[b].\size+root.blocks[end].\size}. Beause otherwise, then \nf{(end-c)/2} satisfied the $\nf{root.blocks[(end-c)/2].sum\sub{enq}}<\nf{e}$. In line \ref{doubling} the differnece between \nf{end} and \nf{start} is doubled. See Figure \ref{fig::doubling}.

 After computing \nf{b}, the value \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}). So the routine non constant part is the binary search which takes $\Theta(\log$\nf{root.blocks[b].\size+root.blocks[end].\size)}$)$ steps from the first paragraph.

>>>>

\end{multicols}
\end{algorithmic}
\end{algorithm}
\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
\cmt{Returns \tt{<b,i>} if $E_e(root)=E_i(root,b)$.}
\State \tt{start= end-1} \label{dsearchStart}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start= max(start-(end-start), 0)} \label{doubling}
\EndWhile
\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)} \label{dsearchEnd}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
\State\Return \tt{<b,i>}
\EndFunction{DSearch}
\end{algorithmic}
\end{algorithm}

@@

\begin{lemma}[DSearch correctness] \label{dsearch}
If $\nf{root.blocks[end].sum\sub{enq}} \geq \nf{e}$, \nf{DSearch(e, end)} returns \nf{<b, i>} such that $E_i(root,b)= E_e(root)$.
\end{lemma}
\begin{proof}
\nf{DSearch} performs a doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} to find $E_e(root)$. From Lemma we know \nf{sum\sub{enq}} fields of nf{root.blocks[]} are sorted in a non-decreasing order. Since \nf{root.blocks[0].sum\sub{enq}=0} and there is a \nf{block} in the root with \nf{sum\sub{enq}} value greater than \nf{e}, so there is a \nf{b} that $\nf{root.blocks[b].sum\sub{enq}} \geq e$ but $\nf{root.blocks[b-1].sum\sub{enq}} < e$. This block contains $E_i(root,b)$ and  the search on Line \ref{dsearchStart}-\ref{dsearchEnd} will eventually reach the \nf{b}.

\end{proof}
\begin{lemma}[DSearch Analysis]\label{dsearchTime}
  Assume $\nf{root.blocks[end].sum\sub{enq}} \geq \nf{e}$ and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}, then  \nf{DSearch(e, end)} takes $\Theta(\log$\nf{root.blocks[b].\size+ root.blocks[end].\size}$)$ steps.
\end{lemma}
\begin{proof}
First we show $\nf{end} - \nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1 \big)$. From line \ref{addOP}, we know that \nf{num} field of the every block in the tree is greater than 0. So, each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue} or at least one \nf{Dequeue}. Suppose there were more than \nf{root.blocks[b].\size}\nf{Dequeue}s in \nf{root.blocks[b+1..end-1]}. Then the element in the queue which is the response to the \nf{Dequeue()} would become dequeued at some point after \nf{blocks[b]}'s last operations and before \nf{root.blocks[end]}'s first operation. Which means the response to to a \nf{Dequeue} in \nf{root.blocks[end]} could not be in $E(n,b)$. Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the operations of \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end-1].\size} \nf{Enqueue}s in \nf{root.blocks[b+1..end-1]}. Because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size of the queue is \nf{root.blocks[end-1].\size}. Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}. So $\nf{end-b}\leq 2 \times\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1$. See Figure \ref{fig::doubling}.

Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]} then with doubling search in $\Theta \big(\log($\nf{root.blocks[b].\size+root.blocks[end].\size}$)\big)$ steps we reach \nf{start=c} that the \nf{root.blocks[c].sum\sub{enq}} is less than \nf{e} and \nf{end-c} is not more than $2 \times 2 \times \big(\nf{root.blocks[b].\size+root.blocks[end].\size}\big)$. Beause otherwise, then \nf{(end-c)/2} satisfied the $\nf{root.blocks[(end-c)/2].sum\sub{enq}}<\nf{e}$. In line \ref{doubling} the difference between \nf{end} and \nf{start} is doubled. See Figure \ref{fig::doubling}.

 After computing \nf{b}, the value \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}). So the whole \nf{DSearch} routine takes $\Theta\big(\log(\nf{root.blocks[b].\size+root.blocks[end].\size})\big)$ steps.
